### 2014-09-10 FlyingPig-journal/conclusions.tex
New space for reviewing the methodology paper and updates for the SM paper

This paper presented \pisodm for designing and developing reliable service-based applications. 
\pisodm is an MDA methodology that extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
These requirements are taken into account from the initial stages of the software development process.
Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.

>>>>

This paper presented \pisodm, an MDA methodology for designing and developing reliable service-based applications. 
\pisodm extends a previously defined method (called SOD-M) to include Non-Functional Requirements.
These requirements are taken into account from the early stages of the software development process.
Non-functional constraints are related to business rules associated to the behavior of the application and, in the case of service-based applications, they are also concerned with constraints imposed by the services.

### 2014-09-10 FlyingPig-journal/intro.tex
New space for reviewing the methodology paper and updates for the SM paper

Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
%
%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather, 
%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed. 
Ideally, non-functional requirements should be considered in every phase of the software development.
Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
In consequence, the development process will not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.

>>>>

Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
%
%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather, 
%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed. 
Ideally, non-functional requirements should be considered in every phase of the software development.
Yet,  they are partially or rarely methodologically derived from the specification, being usually added once the code has been implemented. 
In consequence, the development process does not fully preserve the compliance and reuse expectations provided by the service oriented computing methods.

### 2014-09-10 FlyingPig-journal/modelingCrimes.tex
New space for reviewing the methodology paper and updates for the SM paper

>>>>

In order to introduce the context of our work consider the Tracking crime application where civil population and by police share information about criminality in given zones of an imaginary city. 
Users signal crimes using twitter and police officers notify crimes  they have to deal with. 
Some of this information, if it is not confidential, can be shared to the community of users using this application. 
Users can track crimes in given zones and see information located in a map according to their privileges.  
For example civilians can ask Locate the crimes done the last month 100 km from my current position that happened between 8:00 and 14:00.
While police officers can ask Locate the regions in my sector where murders happened in the last month
This information can come from the police database and from twitter posts. 
The zones of the city are thereby according to their degree of criminality.

 In order to provide these functions the application benefits from existing services that provide information,  storage and data visualization functions. Thus, the application is a service based application that implements a business process.
 The business process is specified in terms of ordered tasks and define the application logic of a Service oriented Application. Tasks can be performed by a person or an entity. In our context, tasks are implemented by Services. 
Service is an application implemented by a provider that exports an API through a network (e.g., Internet)
API is defined using an interface definition language (IDL, e.g., WSDL for Web services).
 In our example the business procces can start with one of two tasks: Notify a crime, or track a crime. A notified crime can be then stored in a database. Tracked crimes are visualized in a map and then the used can ask for detailed information. They are implemented by four services: twitter and an adhoc police service for notifying crimes, Amazon as persistence service and google maps for visualizing and locating crimes in a map. 
Note that a task can be implemented by a service composition. This is the case of the task Notify crime in our example that enables to notify crimes through twitter or through the police notification service.

Business processes have also associated rules and constraints that define their non functional requirements.  
NFR represents the “semantics” and the conditions in which the tasks must be done. 
In our example we have some constraints. 
\begin{itemize}
\item Twtter requires to be accessed through an authentication protocol, and the police crime notification service has a control  access protocol and allows only 3 tries to access it. 
\item Only users with enough privileges can store crimes’ notifications. 
\item Users can only track crimes notified by providers t that hey have authorization to contact; For example civil population cannot track all the crimes notified by the police. 
\item If the google map is unavailable the results of a track request are delivered on text. 
\item If a user tracks crimes that are provided by a service for which she does not have authorization then show an empty map. 
\item Detailed information about a crime depends on the user privileges, on the police assignment regions. 
\item The tracking crime app keeps track of the accesses to detailed information about crimes. 
\end{itemize}

Through the example we underlined that every application implements functional aspects that describe its application logic. Recall that an application logic refers to routines that perform the activities to reach the application objective.
Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.

The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
API does not necessarily export information about methods dependency (e.g., in the REST protocol);
they do not share their state (stateless).

### 2014-09-10 FlyingPig-journal/modelingFlyingPig.tex
New space for reviewing the methodology paper and updates for the SM paper

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
%\subsection{Platform-Specific Models}
%
%
%This level focusses on the functionality, in the context of a particular implementation infrastructure.
%Models at this level combine the platform-independent view with specific aspects of the application to be deployed. 
%At the PSM level we have lower-level models that can be automatically translated into any service composition language. 
%We have defined one meta-model at this level, called \textit{$\pi$-PEWS}.
%
%The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions, including non-functional features.
%\textit{$\pi$-PEWS} models have two main parts:
%\begin{itemizedTrivlist}
%\item \textit{Service compositions (workflows)} formed by individual service operations and composite operations defined by the use of workflow operators, such as guarded operations ($[C]S$); sequential ($\ . \ $), parallel ($\ \| \ $) and alternative ($\ + \ $) compositions; as well as sequential ($*$) and parallel ($\{\dots\}$) repetitions.
%
%\item \textit{ECA Rules}, with a classic semantics, i.e, {\em when an event of type E occurs, if condition C is verified then execute the action A}. 
%These rules are used to specify non-functional behaviour.
%\end{itemizedTrivlist}
%
%The workflow constructors of \textit{$\pi$-PEWS} are similar to those of  most service orchestration languages, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
%
%
%%%
%%\begin{figure}[t]
%%\centering
%%\includegraphics[width=1.0\textwidth]{figs/PEWSMetamodel}
%%\caption{$\pi$-{\sc Pews} Metamodel}
%%\label{fig:metamodel}
%%\end{figure}
%%%
%
%\bigskip
%The following example presents the \textit{$\pi$-PEWS} model for the ``To Publish Music'' application.
%\begin{example}[To Publish Music (end)]\label{Colocar-o-label-correto-aqui}
%The next code presents the \textit{$\pi$-PEWS}  model for our running example.\footnote{\color{red} Placido: Please change the pdf file as needed!}
%
%\begin{center}
%\includegraphics[scale=0.7]{figs/pewsEspecificationContract.pdf}
%\end{center}
%
%The initial part of the model consists of a workflow expression.
%In the given code, the execution of \texttt{searchSong} is followed by \texttt{selectSong}.
%At this point, the user may either buy, download and listen a song or just listen to a song (conditional workflow).
%The last part of the workflow publishes the title of the song either on tweeter or facebook.
%
%ECA rules are specified as contracts~\cite{Meyer92}.
%In our case, the \texttt{buySongContract} specifies conditions to be verified upon the execution of \texttt{buySong}.
%This contract specifies pre and post-conditions, as well as time constraints.\footnote{\color{red} Placido: Please verify the pre-condition assertions!}
%The precondition (EXPLAIN!)\dots
%
%In case of failure in verifying the precondition, the service \texttt{buySong} is retried.
%
%The postcondition ensures that the payment succeeds before .
%
%The time constraint indicates that the \texttt{sendConfirmation} service is invoked immediately after \texttt{buySong}.
%\end{example}
%
%
%
%
%Given a $\pi$-SCM model of a specific service based application (expressed according to the $\pi$-SCM meta-model), it is possible to generate its corresponding $\pi$-{\sc Pews} model. 
%The following section describes the transformation rules between the $\pi$-SCM and $\pi$-{\sc Pews} meta-models.
%
%
%
%
%
%
%
%The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
%provides meta-classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
%The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model meta-classes.} and the {\sc Actions} they perform. 
%Instances of this meta-model are represented as UML activity diagrams. 
%In Figure~\ref{fig:e-scomposition-metamodel}  coloured boxes illustrate classes  modelling  non-functional properties and 
% white boxes represent classes modelling functional ones. 
%
%%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
%\begin{itemizedTrivlist}
%\item A {\sc Business Collaborator} meta-class represents the classes of entities that collaborate in  business processes by performing some  required action. 
%An instance of this meta-class is graphically represented as a partition in the activity diagram. 
%A collaborator can be either internal or external to the system. 
%When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.
%
%\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class Action. 
%A class action represents some type of transformation or processing. 
%There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).
%\begin{figure}[t]
%\centering
%\includegraphics[width=1.0\textwidth]{figs/E-service-composition-metamodel}
%\caption{$\pi$-Service Composition ($\pi$-SCM) Metamodel.}
%\label{fig:e-scomposition-metamodel}
%\end{figure}
%
%\item The {\sc ServiceActivity} meta-class represents classes of composite activity types that must be carried out as part of a business service and is composed by one or more executable nodes.
%
%\item In order to represent constraints types associated to services compositions, we extended introduced the meta-classes {\sc Rule} and {\sc A-policy} (see blue meta-classes in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
%We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
%An {\em A-policy} is defined by attributes and rules. 
%Intuitively, the conditions of each rule will be checked.
%In case of no compliance, the actions defined by the rule will be performed.
%The {\sc Rule} meta-class represents the types of event - condition - action rules where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
%An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
%\end{itemizedTrivlist}
%
%\begin{figure}[t]%[htpb]
%\centering
%\includegraphics[width=0.95\textwidth]{figs/e-composition-model}
%
%{\color{red}\LARGE PLACIDO: Please change the names of the boxes in accordance to the explanation --Martin}
%
%\caption{$\pi$-SCM for the ``To publish music'' business service.}
%\label{fig:servicecompositionmodel}
%\end{figure}
%
%\begin{example}[To Publish Music]\label{ex:toPublicMusic}
%To illustrate the use of the $\pi$-Service Composition Meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}). 
%In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
%% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). 
%The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}. 
%Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
%Both {\em Facebook} and {\em Twitter} services require authentication protocols in order to execute methods that will read and update the user space. 
%%A call to such services must be part of the authentication protocol required by these services.
%In the example we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, that will be associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}). 
%In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol will be associated to the activity {\sf\small UpdateFacebook}.
%
%{\sf\small OAuthPolicy} will implement the open authentication protocol.
%The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token} that will be used to store the authentication token provided by the service.
%This variable is imported through the library {\sf\small OAuthPolicy.Token}. 
%The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: (R$_1$): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and (R$_2$): if the token has expired, then it is renewed. 
%Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
%
%{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol. 
%The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.  
%The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}. 
%On the notification of an event of that type, a credential is obtained using the username and password. 
%\hfill\openbox
%\end{example}
%
%We propose the use of rules and policies to model and associate non-functional properties to service compositions.
%These artifacts will be used to generate the actual programs that will implement the application:
%Once the $\pi$-Service Composition Model has been defined, then it can be transformed into a lower level model (in our case, $\pi$-PEWS) that gives support to code generation. 
%The $\pi$-PEWS  meta-model is described in the next section.

@@

Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
%\footnote{\color{red} This sentence need to move to section 3.1!!}
We identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
Brokers emit requests for risk assessment for one or more companies. 
ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
The values exchanged between ORCA and the brokers are: 
(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
(ii) \textsl{Amendments} which are ORCA's recommendations, based on the data provided by the answers to questionnaires; 
(iii) \textsl{Evaluation Reports} for the client companies; 
and 
(iv) the risk assessment \textsl{Fee}.

The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emited by a particular company. 
Once this need has been declared, the value exchanges between ORCA and Broker are triggered. 
The client company will provide ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
ORCA will provide amendments (recommendations to change practices) and an evaluation report.

@@

The risk assessment process starts after a request from a company.
This corresponds to the value model, in which the start stimulus triggers the whole process.
The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
Questionnaires are considered tasks that users will have to perform. 
Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecure access to the premises of the company. They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.

Once tasks have been completed, they are stored and analyzed to generate a list of noncompliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
During the process of analyzing a questionnaire, the answers of some questions can trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.

Business processes also have associated rules and constraints to define their non-functional requirements (NFR):
\begin{numtrivlist}
\item An acknowledgement is due less than 30 seconds after registering a task or demand for assessment. 
\item The system should be able to deal with, at least, 200 users. 
\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests. 
\item The proviledges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
\item The privileges of Users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
\item All questionnaires need to be fully answered, in order to consider that a task is completed.
\item There is a time limit (in days) for each amendment required by the system.
\end{numtrivlist}

@@

Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
%Use cases 
that describe the main functionalities of the system.
The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.

In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as coloured ovals in Figure~\ref{fig:piUseCaseModel}). 
The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.

@@

The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
Use cases are transformed into \textit{actions} and are represented by white boxes.
The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by coloured boxes) and may be decorated with pre- and post-conditions. 
We can see that this model refines the concepts defined in the $\pi$-UseCase model.
The assertions specify those non-functional requirements, as they are seen by the actors. 
The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.

@@

The model presented in Figure~\ref{fig:PiServiceCompositionModel} 
%offers a system-centric view of the application.
%The previous model 
is now  completed with the services that will provide some functions of the application  \FlyingPig\.
The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
These policies express the pre- and post-conditions of the previous model.
They are associated to the actions of the system.

>>>>

@@

Figure~\ref{fig:E3valuemodel} shows the value model for the \FlyingPig\ application.
It is a business model that graphically represents a business case as a set of value exchanges ($\triangleright$ and $\triangleleft$) and value activities (rounded boxes) performed by business actors (squared boxes).
%\footnote{\color{red} This sentence need to move to section 3.1!!}
We identify two business actors: \textsl{ORCA} and \textsl{Broker}. 
Brokers emit requests for risk assessment for one or more companies. 
ORCA has two value activities which are services that provide an economical benefit:  \textsl{Identify Amendments} and \textsl{Assess Risk Situation}. 
The values exchanged between ORCA and the brokers are: 
(i) \textsl{Questionnaire and Evidences} filled with information about the client company;
(ii) \textsl{Amendments} which are ORCA's recommendations, based on the data provided by the answers to questionnaires; 
(iii) \textsl{Evaluation Reports} for the client companies; 
and 
(iv) the risk assessment \textsl{fee}.

The dependency path in Figure~\ref{fig:E3valuemodel} initiates with the need of assessment emitted by a particular company. 
Once this need has been declared, the value exchanges between ORCA and Broker are triggered. 
The client company provides ORCA with information (answers to a questionnaire), evidence (to support the information) and a fee (monetary value).
Then, ORCA suggests amendments (recommendations to change practices) and provides an evaluation report.

@@

The risk assessment process starts after a request from a company.
This corresponds to the value model, in which the start stimulus triggers the whole process.
The request leads to the definition of a group of users that will answer questionnaires for evaluating risk.
Questionnaires are considered tasks to be performed by users. 
Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company. 
They can be also abstract  such as the protocol used for accessing data on the company's computer server.}.

Once tasks have been completed, they are stored and analyzed to generate a list of non-compliant situations, associated to their corresponding \textit{calls for amendment} (if needed) or a report specifying a compliance level, incidents and a risk map.
During the process of analyzing a questionnaire, the answers to some questions may trigger the generation of additional questionnaires or amendments, that will be scheduled as new tasks.

Business processes also have associated rules and constraints to define their non-functional requirements (NFR):
\begin{numtrivlist}
\item An acknowledgement is due in less than 30 seconds after registering a task or demand for assessment. 
\item The system should be able to deal with, at least, 200 users. 
\item If the number of requests exceeds 200, \FlyingPig\ should implement a load balance strategy for processing the requests. 
\item The privileges of the Channel-Broker must be verified \textit{before} the execution of the actions associated to the \textsf{designate user in charge} $\pi$-use case.
\item The privileges of users must be verified \textit{before} the execution of the actions associated to the \textsf{answer questionnaire and add evidences} $\pi$-use case.
\item All questionnaires need to be fully answered, in order to consider that a task is completed.
\item There is a time limit (in days) for each amendment required by the system.
\end{numtrivlist}

@@

Each actor is associated to one or more $\pi$-use cases (depicted as white ovals in Figure~\ref{fig:piUseCaseModel}). 
%Use cases 
that describe the main functionality of the system.
The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-use cases.

In our model, each $\pi$-use case may be associated to one or more (non-functional) constraints (depicted as colored ovals in Figure~\ref{fig:piUseCaseModel}). 
The model defines three types of constraints: \textit{value}, \textit{business} or \textit{exceptional behavior}. 
Each constraint is identified by the word $<<$\textsf{constraint}$>>$ followed by its type.

@@

The \textsf{Company}, \textsf{Broker-Channel} and \textsf{User} actors are transformed into lanes that represent the business collaborators.
Use cases are transformed into \textit{actions} and are represented by white boxes.
The restrictions associated to each $\pi$-use case are transformed into \textit{assertions} (represented by colored boxes) and may be decorated with pre- and post-conditions. 
We can see that this model refines the concepts defined in the $\pi$-UseCase model.
The assertions specify those non-functional requirements, as they are seen by the actors. 
The next step in the development is to add these assertions to the models that specify the \FlyingPig\ system.

@@

The model presented in Figure~\ref{fig:PiServiceCompositionModel} 
%offers a system-centric view of the application.
%The previous model 
is now  completed with the services that provides functions of the application  \FlyingPig\.
The assertions in Figure~\ref{fig:PiServiceProcessModel} are implemented as \textit{policies}.
These policies express the pre- and post-conditions of the previous model.
They are associated to the actions of the system.

### 2014-09-10 FlyingPig-journal/relworks.tex
New space for reviewing the methodology paper and updates for the SM paper

While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is preformed.  
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called Non-functional Properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioural requirements~\cite{Chung91,MylopoulosBook99,Chung2009}. 
Most research on NFRs focusses on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
In the case of service-based applications, non-functional requirements concern the application itself as well as its component services.

@@

D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}. 
\textit{Quality dimensions} are used to quantify an individual characteristic.
For instance, the quality category \textit{performance} groups characteristics such as
\textit{latency} and \textit{throughput}. 
The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services. A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties as \textit{availability}, \textit{reliability} and \textit{access control}.

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolset for specifying and implementing web service compositions with support to several NFRs. 
The approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
Non-functional concern is a general term used to describe non-functional requirements (such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}). 
Each concern is refined into a set of non-functional attributes.  
Each non-functional attribute represents some behavior, to be refined into non-functional actions. 
An example of non-functional action is \textit{encryption}, which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern. 
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

@@

%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications. 
%The work  presents a development methodology, although this methodology is not centered on NFRs.
The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development processes that take NFRs into account. 
In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modelling is used for
system specification, including NFRs. 
Most authors focus their work on modelling service compositions, although a significant number of approaches is focused on the definition of NFR models.

In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be 
annotated with functional and non-functional attributes (NFAs). 
Functional and non-functional attributes are independently defined.
They are attached to specific tasks later in the development of the application. 
NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.

@@

\bigskip
Most works agree on distinguishing three points of view, namely the point of view of the organization (or \textit{Business} view), of the individual service providers (or \textit{Service} view) and of the composition designer (or \textit{System} view).
The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
Service and System views are concerned with the implementation of the software solution: The Service level is concerned with the building blocks of the application.
It may use web services provided by third party sources.
The System level is concerned with the coordination of services, to implement the business logic.

>>>>

While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is preformed.  
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called Non-functional Properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}. 
Most research on NFRs focusses on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}. 
In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.

@@

D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}. 
\textit{Quality dimensions} are used to quantify an individual characteristic.
For instance, the quality category \textit{performance} groups characteristics such as
\textit{latency} and \textit{throughput}. 
The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services. A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs. 
The approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
Non-functional concern is a general term used to describe non-functional requirements (such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}). 
Each concern is refined into a set of non-functional attributes.  
Each non-functional attribute represents some behavior, to be refined into non-functional actions. 
An example of non-functional action is \textit{encryption}, which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern. 
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

@@

%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications. 
%The work  presents a development methodology, although this methodology is not centered on NFRs.
The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development processes that take NFRs into account. 
In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs, 
whereas in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
system specification, including NFRs. 
In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be 
annotated with functional and non-functional attributes (NFAs). 
Functional and non-functional attributes are independently defined.
They are attached to specific tasks later in the development of the application. 
NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.

@@

\bigskip
Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
It may use web services provided by third party sources.
The System level deals with the coordination of services, to implement the business logic.
Our proposal, shown in the next section, follows these ideas.
